# Data Visualizations

## How to Read these Plots {-}

The following plots display a lot of data in a visually-dense format. The indicators for each COO Community are arranged in a column, with each grid square containing the following information:

  * the indicator values for each census tract (represented by a thin yellow line)
  * the indicator values for all census tracts (represented by a light gray histogram chart in the grid's background)
  * the indicator value for King County (represented by a thick blue line)
  * _for percent indicators only:_ the indicator value of all of the tracts in the COO community combined (represented by a thick dashed vertical line)
  
Reading the plot from top to bottom, one column at a time, the viewer can see how the tracts of each community compare to each other; to the community as a whole; to the county as a whole; and to the distribution of values amongst all King County census tracts.

## Plots {-}

```{r app-data-viz-comm-vuln-hist, fig.asp = 1, fig.cap='Vulnerability Indicators by COO Community (2013-2017)'}

# SETUP -------------------------------------------------------------------

source("R/setup.R",echo = FALSE)

# LOAD DATA ---------------------------------------------------------------

coorev18 <- readr::read_rds("data/model-all-20190305.rds") %>% 
  select(matches("GEOGRAPHY"), matches("COOREV18")) %>%  
  as_tibble() 


# TRANSFORM DATA: VULNERABILITY -------------------------------------------


dat_vuln_wide <- coorev18 %>% select(matches("GEOGRAPHY"),
                 COOREV18_VULN_POC_EST_PCT_2013_2017,
                 COOREV18_VULN_LOWED_EST_PCT_2013_2017, 
                         COOREV18_VULN_LOWINCAMI_EST_PCT_2011_2015,
                 COOREV18_VULN_RENTHH_EST_PCT_2013_2017) 

dat_vuln_long <- dat_vuln_wide %>% 
  gather(COL, VAL, -matches("GEOGRAPHY")) %>% 
  mutate(YEAR = str_extract(COL,"\\d{4}_\\d{4}$")) %>% 
  mutate(VAR = case_when(
    str_detect(COL,"LOWED") ~ "% LOW-EDUCATION",
    str_detect(COL,"LOWINCAMI") ~ "% LOW-INCOME",
    str_detect(COL,"POC") ~ "% PEOPLE OF COLOR",
    str_detect(COL,"RENTHH") ~ "% RENTERS"
  )) %>% 
  mutate(VAR = glue("{VAR} ({str_replace(YEAR,'_','-')})"))

dat_vuln_tr <- dat_vuln_long %>% 
  filter(GEOGRAPHY_TYPE %in% c("tract"))

dat_vuln_lines <- dat_vuln_long %>% 
  filter(GEOGRAPHY_TYPE %in% c("county", "community") | !is.na(GEOGRAPHY_COMMUNITY_NAME)) %>% 
  mutate(GEOG_TYPE_FCT = factor(GEOGRAPHY_TYPE,
                                levels = c("tract",
                                           "community",
                                           "county")))
dat_vuln_lines_by_comm <- dat_vuln_lines %>% 
  mutate(WC = VAL,
         RV = VAL,
         STC_TUK = VAL) %>% 
  select(-VAL) %>% 
  gather(GEOG_COMM, VAL, WC, RV, STC_TUK) %>% 
  nest(-GEOG_COMM) %>% 
  mutate(data = map2(data, GEOG_COMM, ~ filter(.x, GEOGRAPHY_COMMUNITY_ID %in% .y | GEOGRAPHY_TYPE %in% c("county", "community")))) %>% 
  unnest() %>% 
  mutate(GEOG_COMM = case_when(
    GEOG_COMM %in% "WC" ~ "White Center",
    GEOG_COMM %in% "RV" ~ "Rainier Valley",
    GEOG_COMM %in% "STC_TUK" ~ "Seatac/Tukwila"
  )) %>% 
  filter(GEOGRAPHY_TYPE %in% c("tract","county") | GEOG_COMM == GEOGRAPHY_NAME)


# PLOT: VULNERABILITY ------------------------------------------------------

showtext_auto()

show_comm_hist <- function(x, y_axis_title, legend_position){  

  comm_name <- unique(x$GEOG_COMM)
  
  dat_vuln_tr %>% 
    ggplot2::ggplot(ggplot2::aes(x = VAL)) + 
    ggplot2::geom_histogram(fill = colors_nct$grey) +
    ggplot2::geom_vline(data = x, 
                        aes(xintercept = VAL, 
                            color = GEOG_TYPE_FCT, 
                            size = GEOG_TYPE_FCT, 
                            alpha = GEOG_TYPE_FCT,
                            linetype = GEOG_TYPE_FCT)) +
    scale_color_manual(values = c(colors_nct_list$'Early Type 1',
                                  colors_nct_list$'Early Type 1',
                                  colors_nct_list$Late)) +
    scale_x_continuous(limits = c(0,1), labels = scales::percent) +
    scale_size_manual(values = c(0.5,1,1)) +
    scale_alpha_manual(values = c(0.75,1,1)) +
    scale_linetype_manual(values = c(1,6,1)) + 
    ggplot2::facet_wrap(~ VAR, ncol = 1, scales = "free_x") +
    theme_nct_8 +
    labs(y = y_axis_title,
         x = "",
         title = comm_name) +
    theme(legend.position = legend_position,
          legend.direction="horizontal",
          legend.text = element_text(hjust = 0.5),
          legend.spacing.x = unit(2, "pt"),
          legend.key.width = unit(30, "pt"),
          legend.key.height = unit(12, "pt"),
          legend.title = element_blank(),
          strip.background = element_blank(), 
          strip.placement = "outside")
}
dat_vuln_list_by_comm <- group_split(dat_vuln_lines_by_comm, GEOG_COMM)

pmap(list(x = dat_vuln_list_by_comm,
            y_axis_title = c("# of Census Tracts","",""),
            legend_position = c("none","bottom","none")), 
     show_comm_hist) %>% 
  wrap_plots(nrow = 1) +
  plot_annotation(title = "Vulnerability Indicators by COO Community (2013-2017)",
                subtitle = "",
                caption = wrapper("Source: author's analysis of data from American Community Survey and HUD CHAS
                                  \nNote: The income indicator uses CHAS data, which lags behind the ACS data by two years. The gray bars show the amount of census tracts at each part of the x-axis."),
                theme = theme_nct)

```


```{r app-data-viz-comm-demo-hist, fig.asp = 1.25, fig.cap='Demographic Change Indicators by COO Community (2007-2001 to 2013-2017)'}

# TRANSFORM DATA: DEMO. CHANGE -------------------------------------------


dat_demochng_wide <- coorev18 %>% select(matches("GEOGRAPHY"),
                                         COOREV18_DEMOCHNG_HIGHED_DIFF_PCT_2007_2011_TO_2013_2017,
                                         COOREV18_DEMOCHNG_INCOME_DIFF_MED_2007_2011_TO_2013_2017,
                                         COOREV18_DEMOCHNG_WHITE_DIFF_PCT_2007_2011_TO_2013_2017,
                                         COOREV18_DEMOCHNG_OWNHH_DIFF_PCT_2007_2011_TO_2013_2017) 

dat_demochng_long <- dat_demochng_wide %>% 
  gather(COL, VAL, -matches("GEOGRAPHY")) %>% 
  mutate(YEAR = str_extract(COL,"\\d{4}_\\d{4}_TO_\\d{4}_\\d{4}$")) %>%  
  mutate(YEAR = str_replace(YEAR,"_TO_", " TO ")) %>% 
  mutate(YEAR = str_replace_all(YEAR,'_','-')) %>% 
  mutate(YEAR = str_replace_all(YEAR,'20',"'")) %>% 
  mutate(VAR = case_when(
    str_detect(COL,"HIGHED") ~ "CHANGE % HIGH-ED.",
    str_detect(COL,"INCOME") ~ "CHANGE MEDIAN INC.",
    str_detect(COL,"WHITE") ~ "CHANGE % WHITE",
    str_detect(COL,"OWNHH") ~ "CHANGE % HOMEOWNER"
  )) %>% 
  mutate(VAR = as.character(glue("{VAR} ({YEAR})"))) 

lvls <- unique(dat_demochng_long$VAR)

dat_demochng_long <-dat_demochng_long %>% 
  mutate(VAR = factor(VAR,
                      levels = c(lvls)))


dat_demochng_tr <- dat_demochng_long %>% 
  filter(GEOGRAPHY_TYPE %in% c("tract"))

dat_demochng_lines <- dat_demochng_long %>% 
  filter(GEOGRAPHY_TYPE %in% c("county", "community") | !is.na(GEOGRAPHY_COMMUNITY_NAME)) %>% 
  mutate(GEOG_TYPE_FCT = factor(GEOGRAPHY_TYPE,
                                levels = c("tract",
                                           "community",
                                           "county")))
dat_demochng_lines_by_comm <- dat_demochng_lines %>% 
  mutate(WC = VAL,
         RV = VAL,
         STC_TUK = VAL) %>% 
  select(-VAL) %>% 
  gather(GEOG_COMM, VAL, WC, RV, STC_TUK) %>% 
  nest(-GEOG_COMM) %>% 
  mutate(data = map2(data, GEOG_COMM, ~ filter(.x, GEOGRAPHY_COMMUNITY_ID %in% .y | GEOGRAPHY_TYPE %in% c("county", "community")))) %>% 
  unnest() %>% 
  mutate(GEOG_COMM = case_when(
    GEOG_COMM %in% "WC" ~ "White Center",
    GEOG_COMM %in% "RV" ~ "Rainier Valley",
    GEOG_COMM %in% "STC_TUK" ~ "Seatac/Tukwila"
  )) %>% 
  filter(GEOGRAPHY_TYPE %in% c("tract","county") | GEOG_COMM == GEOGRAPHY_NAME)


# PLOT: DEMOGRAPHIC CHANGE ------------------------------------------------

showtext_auto()

show_comm_hist <- function(x, y_axis_title, legend_position){  
  
  comm_name <- unique(x$GEOG_COMM)
  
  varied_labels <- function(x) {   
    
    if (max(x, na.rm = TRUE) > 1){
      percent_format(accuracy = 1,prefix = "+")(x) %>% str_replace_all("\\+\\-","\\-")
    }else{
      percent_format(accuracy = 1, prefix = "+", suffix = "%pts.")(x) %>% str_replace_all("\\+\\-","\\-")
    } 
  }
  
  dat_demochng_tr %>% 
    filter(!is.na(VAL)) %>% 
    ggplot2::ggplot(ggplot2::aes(x = VAL)) + 
    ggplot2::geom_histogram(fill = colors_nct$grey) +
    ggplot2::geom_vline(data = x, 
                        aes(xintercept = VAL, 
                            color = GEOG_TYPE_FCT, 
                            size = GEOG_TYPE_FCT, 
                            alpha = GEOG_TYPE_FCT,
                            linetype = GEOG_TYPE_FCT)) +
    scale_color_manual(values = c(colors_nct_list$Dynamic,
                                  colors_nct_list$Dynamic,
                                  colors_nct_list$Late)) + 
    scale_size_manual(values = c(0.5,1,1)) +
    scale_alpha_manual(values = c(.75,1,1)) +
    scale_linetype_manual(values = c(1,6,1)) +  
     facet_wrap(~ VAR, ncol = 1, scales = "free_x") +
    scale_x_continuous(labels = varied_labels) +
    theme_nct_8 +
    labs(y = y_axis_title,
         x = "",
         title = comm_name) +
    theme(legend.position = legend_position,
          legend.direction="horizontal",
          legend.text = element_text(hjust = 0.5),
          legend.spacing.x = unit(2, "pt"),
          legend.key.width = unit(30, "pt"),
          legend.key.height = unit(12, "pt"),
          legend.title = element_blank(),
          strip.background = element_blank(), 
          strip.placement = "outside")
}
dat_demochng_list_by_comm <- group_split(dat_demochng_lines_by_comm, GEOG_COMM)

pmap(list(x = dat_demochng_list_by_comm,
          y_axis_title = c("# of Census Tracts","",""),
          legend_position = c("none","bottom","none")), 
     show_comm_hist) %>% 
  wrap_plots(nrow = 1) +
  plot_annotation(title = "Demographic Change Indicators by COO Community (2007-2001 to 2013-2017)",
                  subtitle = "",
                  caption = wrapper("Source: author's analysis of data from American Community Survey 
                                  \nNote: Median income at the community level is not calculated. The gray bars show the amount of census tracts at each part of the x-axis."),
                  theme = theme_nct)



```


```{r app-data-viz-comm-hous-hist, fig.asp = 1.5, fig.cap='Housing Market Indicators by COO Community (2018)'}
# TRANSFORM DATA: HOUSING MARKET -------------------------------------------


dat_housmrkt_wide <- coorev18 %>% select(matches("GEOGRAPHY"),
                                         COOREV18_HOUSMRKT_RENTPRICE_EST_MED_2013_2017,
                                         COOREV18_HOUSMRKT_SFHOMEVALKC_EST_MED_2018_2018,
                                         COOREV18_HOUSMRKT_SFSALEPRICESQFT_EST_MED_2018_2018,
                                         COOREV18_HOUSMRKT_SFHOMEVALKC_DIFF_MED_2010_2010_TO_2018_2018,
                                         COOREV18_HOUSMRKT_SFSALEPRICESQFT_DIFF_MED_2010_2010_TO_2018_2018,
                                         COOREV18_HOUSMRKT_SFSALERATE_EST_PCT_2016_2018) 

dat_housmrkt_long <- dat_housmrkt_wide %>% 
  gather(COL, VAL, -matches("GEOGRAPHY")) %>% 
  mutate(YEAR = str_extract(COL,"20.*$")) %>% 
  mutate(YEAR = str_replace_all(YEAR, "2010_2010","2010")) %>% 
  mutate(YEAR = str_replace_all(YEAR, "2018_2018","2018")) %>% 
  mutate(YEAR = str_replace_all(YEAR, "_TO_"," TO ")) %>%  
  mutate(YEAR = str_replace_all(YEAR,'_','-')) %>%  
  mutate(YEAR = str_replace_all(YEAR,'20',"'")) %>% 
  mutate(YEAR = str_replace(YEAR,"^'18$","2018")) %>%  
  mutate(VAR = case_when(
    str_detect(COL,"RENTPRICE") ~ "MEDIAN RENT PRICE",
    str_detect(COL,"SFHOMEVALKC_EST") ~ "MEDIAN HOME VALUE",
    str_detect(COL,"SFSALEPRICESQFT_EST") ~ "MEDIAN HOME PRICE SQFT",
    str_detect(COL,"SFHOMEVALKC_DIFF") ~ "CHANGE MEDIAN HOME VALUE",
    str_detect(COL,"SFSALEPRICESQFT_DIFF") ~ "CHANGE MEDIAN HOME PRICE SQFT",
    str_detect(COL,"SFSALERATE") ~ "% HOMES SOLD"
  )) %>% 
  mutate(VAR = as.character(glue("{VAR} ({YEAR})")))

lvls <- unique(dat_housmrkt_long$VAR)

dat_housmrkt_long <-dat_housmrkt_long %>% 
  mutate(VAR = factor(VAR,
                      levels = c(lvls)))


dat_housmrkt_tr <- dat_housmrkt_long %>% 
  filter(GEOGRAPHY_TYPE %in% c("tract"))

dat_housmrkt_tr_no_outliers <- dat_housmrkt_tr %>% 
  mutate(VAL = case_when(
    VAR %in% "MEDIAN RENT PRICE ('13-'17)" & VAL >= 2500 ~ 2500,
    VAR %in% "MEDIAN HOME VALUE (2018)" & VAL >= 1e6 ~ 1e6,
    VAR %in% "MEDIAN HOME PRICE SQFT (2018)" & VAL >= 800 ~ 800,
    VAR %in% "CHANGE MEDIAN HOME VALUE ('10 TO '18)" & VAL >= 2 ~ 2,
    VAR %in% "% HOMES SOLD ('16-'18)" & VAL >= 0.2 ~ 0.2, 
    TRUE ~ VAL
  ))

dat_housmrkt_lines <- dat_housmrkt_long %>% 
  filter(GEOGRAPHY_TYPE %in% c("county", "community") | !is.na(GEOGRAPHY_COMMUNITY_NAME)) %>% 
  mutate(GEOG_TYPE_FCT = factor(GEOGRAPHY_TYPE,
                                levels = c("tract",
                                           "community",
                                           "county")))
dat_housmrkt_lines_by_comm <- dat_housmrkt_lines %>% 
  mutate(WC = VAL,
         RV = VAL,
         STC_TUK = VAL) %>% 
  select(-VAL) %>% 
  gather(GEOG_COMM, VAL, WC, RV, STC_TUK) %>% 
  nest(-GEOG_COMM) %>% 
  mutate(data = map2(data, GEOG_COMM, ~ filter(.x, GEOGRAPHY_COMMUNITY_ID %in% .y | GEOGRAPHY_TYPE %in% c("county", "community")))) %>% 
  unnest() %>% 
  mutate(GEOG_COMM = case_when(
    GEOG_COMM %in% "WC" ~ "White Center",
    GEOG_COMM %in% "RV" ~ "Rainier Valley",
    GEOG_COMM %in% "STC_TUK" ~ "Seatac/Tukwila"
  )) %>% 
  filter(GEOGRAPHY_TYPE %in% c("tract","county") | GEOG_COMM == GEOGRAPHY_NAME)

# PLOT: HOUSING MARKET ------------------------------------------------

showtext_auto()

show_comm_hist <- function(x, y_axis_title, legend_position){   
  
  comm_name <- unique(x$GEOG_COMM)
  
  dollar_short <- function(x){
    if(is.na(x)){NA_character_} else 
      if(x>=1e6) {dollar_format(accuracy = 1,suffix = "M")(x/1e6)} else
        if(between(x,1e3,1e6-1)) {dollar_format(accuracy = 1,suffix = "k")(x/1e3)} else
          if(between(x,0,1e3-1)){ dollar_format(accuracy = 1,suffix = "")(x)} else 
            stop("Oops")
  }
  
  
  varied_labels <- function(x) {   
    
    if(max(x, na.rm = TRUE) > 1000){
      map_chr(x, dollar_short)
    }else 
      if (max(x, na.rm = TRUE) > 5){
      dollar_format(accuracy = 1, suffix = "/sqft")(x)
    }else
     if(max(x, na.rm = TRUE) > 1){
       percent_format(accuracy = 1,prefix = "+")(x) %>% str_replace_all("\\+\\-","\\-")
     }else{
      percent_format(accuracy = 1, prefix = "", suffix = "%")(x)
    } 
  }
  
  dat_housmrkt_tr_no_outliers %>% 
    filter(!is.na(VAL)) %>% 
    ggplot2::ggplot(ggplot2::aes(x = VAL)) + 
    ggplot2::geom_histogram(fill = colors_nct$grey) +
    ggplot2::geom_vline(data = x, 
                        aes(xintercept = VAL, 
                            color = GEOG_TYPE_FCT, 
                            size = GEOG_TYPE_FCT, 
                            alpha = GEOG_TYPE_FCT,
                            linetype = GEOG_TYPE_FCT)) +
    scale_color_manual(values = c(colors_nct_list$`Early Type 2`,
                                  colors_nct_list$`Early Type 2`,
                                  colors_nct_list$Late)) + 
    scale_size_manual(values = c(0.5,1,1)) +
    scale_alpha_manual(values = c(.75,1,1)) +
    scale_linetype_manual(values = c(1,6,1)) +  
    facet_wrap(~ VAR, ncol = 1, scales = "free_x") +
    scale_x_continuous(labels = varied_labels) +
    theme_nct_8 +
    labs(y = y_axis_title,
         x = "",
         title = comm_name) +
    theme(legend.position = legend_position,
          legend.direction="horizontal",
          legend.text = element_text(hjust = 0.5),
          legend.spacing.x = unit(2, "pt"),
          legend.key.width = unit(30, "pt"),
          legend.key.height = unit(12, "pt"),
          legend.title = element_blank(),
          strip.background = element_blank(), 
          strip.placement = "outside")
}


dat_housmrkt_list_by_comm <- group_split(dat_housmrkt_lines_by_comm, GEOG_COMM)

pmap(list(x = dat_housmrkt_list_by_comm,
          y_axis_title = c("# of Census Tracts","",""),
          legend_position = c("none","bottom","none")), 
     show_comm_hist) %>% 
  wrap_plots(nrow = 1) +
  plot_annotation(title = "Housing Market Indicators by COO Community (2018)",
                  subtitle = "",
                  caption = wrapper("Source: author's analysis of data from the American Community Survey and the King County Tax Assessor's Office. 
                                  \nNote: Some outliers have been transformed to enhance the clarity of the grahic. Median rent price at the community level is not calculated. The gray bars show the amount of census tracts at each part of the x-axis."),
                  theme = theme_nct)





```

